Commutative character sheaves and geometric types for supercuspidal representations

### 2017-06-27 Geometric types.tex
Addressing second round of referee comments

We are extremely grateful to Loren Spice for explaining Yu's types for supercuspidal representations.
We also thank Masoud Kamgarpour for helpful conversations, and the anonymous referee of this paper, would pointed out an error in an earlier version and helped us clarify several points.

@@

We now consider the case of a smooth group scheme without the commutativity assumption.  We start
by relating character sheaves on $G$ to character sheaves on its abelianization.

If $\chi \in G(k)^*$ is a character, it must vanish on the derived subgroup $G\der(k)$. \todo{Error} 
Kamgarpour gives an example \cite{kamgarpour:09a}*{(1.1)} of a character sheaf
on $G = \PGL_n$ that does not vanish on $G\der = \PGL_n$, defined by the extension
\[
1 \to \mu_n \to \SL_n \to \PGL_n \to 1
\]
together with a non-trivial character of $\mu_n(k)$. \todo{Small change here.}
In order to obtain a relationship between character sheaves on $G$ and characters of $G(k)$,
he opts to give a different definition of commutator and, in doing so, introduces a `stacky abelianization' of $G$ in order to geometrize characters of $G(\Fq)$.
 Since we have already seen the need to adapt the
notion of character sheaf, even in the commutative case, we instead add restrictions
to the definition of commutative character sheaf and leave the definition of $G\der$ unchanged, allowing us to use the schematic abelianization of $G$ in the geometrization of characters of $G(\Fq)$.

@@

In order to get character sheaves that correspond to characters in $G(k)^*$, we must discard those
character sheaves whose restriction to the derived subgroup is nontrivial.
Recall from Section~\ref{sec:defs} that we refer to objects in category $\CS(G)$, defined as
in \cite{cunningham-roe:13a}*{Def. 1.1}, as linear character sheaves when $G$ is smooth but
not necessarily commutative.  
We define the following category to track the trivialization on the derived subgroup;
commutative character sheaves will then be defined as a subcategory.

@@

The reason for tracking $\beta$ is that it determines an isomorphism $\gamma : m^*\cs{L} \to \theta^*m^*\cs{L}$ as follows,
which will replace the $\xi$ of Definition \ref{def:CCScom}.
Let $i : G \to G$ be inversion and $c : G\times G\to G\der$ be the commutator map, defined by $c(x,y)= xyx^{-1}y^{-1}$.
Both are smooth morphisms of $\Fq$-schemes.
Set $m' = i \circ m \circ \theta$ and let $j\der : G\der\to G$ be inclusion; then $j\der\circ c = m \circ (m \times m')$.
Then $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ by the diagram of isomorphisms below.
\begin{equation}
\begin{tikzcd}
\arrow[equal]{d} c^* (\cs{L}\vert_{G\der}) \arrow{r}{c^*(\beta)} 
	&  c^*((\EE)_{G\der}) \arrow[equal]{d} \\
\arrow{d}[swap]{j\der\circ c = m \circ (m \times m')} c^* j\der^* \cs{L} 
	&   (\EE)_{G\times G} \\
(m \times m')^* m^* \cs{L} \arrow{d}[swap]{(m \times m')^*(\mu)} 
	&  m^*\cs{L} \otimes \theta^* m^* i^* \cs{L}  \arrow[dashed]{u}[swap]{\gamma'} \\
(m \times m')^* (\cs{L} \boxtimes \cs{L}) \arrow[equal]{r} 
	& m^*\cs{L} \otimes (m')^*\cs{L} \arrow{u}[swap]{m' = i\circ m\circ \theta} 
\end{tikzcd}
\end{equation}
In the diagram above, the arrows labeled with equations come from canonical isomorphisms of functors on Weil sheaves derived from the equations; so, for example, the middle left isomorphism comes from $(m\times m')^* m^* \iso c^* j\der^*$ since $j\der\circ c = m \circ (m \times m')$.
Using the monoidal structure of the category of Weil local systems on $G\times G$, the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ defines an isomorphism
\[
m^*\cs{L} \to (\theta^* m^* i^*\cs{L})^\vee.
\]
Applying the canonical isomorphisms $(\theta^* m^* i^*\cs{L})^\vee \iso \theta^* m^* i^* (\cs{L}^\vee)$ and $i^*(\cs{L}^\vee) \cong \cs{L}$, this map provides the promised isomorphism
\[
\begin{tikzcd}
\gamma : m^*\cs{L} \arrow{r} & \theta^* m^* \cs{L}.
\end{tikzcd}
\]

@@

\begin{lemma}\label{lem:HH}
If $(\cs{L},\mu,\nu)$ is a $G\der$-equivariant linear character sheaf on $G$ then $\mu : m^*\cs{L} \to \cs{L}\boxtimes \cs{L}$ 
is a morphism of $G^2\der$-equivariant Weil local systems on $G^2$. 
\end{lemma}

@@

\begin{corollary}
If $G$ is a smooth group scheme and $(\cs{L},\mu)$ is a character sheaf on $G$ with trivial restriction to $G\der$,
then the set of isomorphism classes of objects in $\CS(G\ab)$ mapping to $(\cs{L},\mu)$ under $q^*$ is a principal homogeneous space
for $C$.
\end{corollary}
\begin{proof}
By Proposition~\ref{prop:CSGabab}, it suffices to find the set of isomorphism classes in $\CSab(G)$ mapping to
$(\cs{L},\mu)$ under the forgetful functor.  By the previous corollary this set is nonempty.
If $(\cs{L},\mu,\beta)$ and $(\cs{L},\mu,\beta')$ both map to $(\cs{L},\mu)$, then $\beta' \circ \beta^{-1}$ is an
automorphism of the constant sheaf on $G\der$.  Conversely, if $\varphi$ is an automorphism of $(\EE)_{G\der}$
and $(\cs{L},\mu,\beta) \in \CSab(G)$ then $(\cs{L},\mu,\varphi \circ \beta) \in \CSab(G)$.  By \cite{cunningham-roe:13a}*{Theorem 3.9},
the automorphism group is isomorphic to $\Hom(\pi_0(\bar{G}\der)_{\Frob{}}, \EEx)$.  Finally, we note that any
automorphism $\alpha$ of $(\cs{L},\mu) \in \CS(G)$ defines an isomorphism $(\cs{L},\mu,\beta \circ \alpha|_{G\der}) \to (\cs{L},\mu,\beta)$.
Applying \cite{cunningham-roe:13a}*{Theorem 3.9} again yields the desired result.
\end{proof}

@@

\begin{lemma}\label{lemma:morinCSab}
If $(\cs{L},\mu,\beta)\in \CS\ab(G)$ then $\mu  : m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$, $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are morphisms in $\CS\ab(G\times G)$.
\end{lemma}
\begin{proof}
Define $m^2 : G^2\times G^2 \to G^2$ by $m^2(g_1,g_2,g_1',g_2') = (g_1g_1',g_2g_2')$. Also define $p^2_i : G^2\times G^2 \to G^2$ by $p^2_i(g_1,g_2,g_1,'g_2') = (g_i,g_i')$.
First we show that $m^*\cs{L}$ is an object in $\CS(G^2)$ by equipping it with an isomorphism $\mu_m^2: (m^2)^* (m^*\cs{L}) \to m^*\cs{L} \boxtimes m^*\cs{L}$ defined by the diagram below.
\[
\begin{tikzcd}
(m^2)^* (m^*\cs{L}) \arrow{d}{(m^2)^*\mu} \arrow[dashed]{rr}{\mu_m^2} && m^*\cs{L} \boxtimes m^*\cs{L}\\
 (m^2)^*(\cs{L}\boxtimes\cs{L}) \arrow[equal]{r} & (m^2)^*(p_1)^*\cs{L} \otimes (m^2)^*(p_2)^*\cs{L} \arrow{r} & (p^2_1)^*m^*\cs{L} \otimes (p^2_2)^*m^*\cs{L} \arrow[equal]{u}
\end{tikzcd}
\] 
The pair $(m^*\cs{L},\mu_m^2)$ satisfies the conditions appearing in \cite{cunningham-roe:13a}*{Def. 1.1}.
The restriction of $m^*\cs{L}$ to $G^2\der= G\der\times G\der$ is canonically isomorphic to $(\EE)_{G^2\der}$ by
\[
\begin{tikzcd}
(m^*\cs{L})\vert_{G^2\der} \arrow{d}{\mu\vert_{G^2\der}} \arrow[dashed]{r}{\beta^2_m} & (\EE)_{G^2\der}\\
(\cs{L}\boxtimes\cs{L})\vert_{G^2\der} \arrow[equal]{r} & (\cs{L}\vert_{G\der})\boxtimes(\cs{L}\vert_{G\der}) \arrow{u}{\beta\boxtimes\beta}.
\end{tikzcd}
\]
This shows that $(m^*\cs{L},\mu_m^2,\beta_m^2) \in \CS\ab(G^2)$. 
Similar work defines $(\cs{L}\boxtimes\cs{L},\mu_\boxtimes^2,\beta_\boxtimes^2) \in \CS\ab(G^2)$.
By construction, $\mu: m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$ is a morphism in $\CS\ab(G^2)$.
Similar work shows that $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are also morphisms in $\CS\ab(G^2)$.
\end{proof}

@@

\begin{proof}
By continuity of $\varphi : \underline{G}(R) \to \EE^\times$, there is some $m \in \NN$ and a factorization
\[
\begin{tikzcd}
\underline{G}(R) \arrow{rr}{\varphi} \arrow{rd} && \EEx\\
& \underline{G}(R/\mathfrak{p}^{m+1}) \arrow{ru}[swap]{\varphi_m} 
\end{tikzcd}
\] 
Set $R_m = R/\mathfrak{p}^{m+1}$ and
set $G_m = \Gr_m^{R}(\underline{G})$, the Greenberg transform of $\underline{G}\times_{\Spec{R}}\Spec{R_m}$.
Then $G_m$ is a smooth group scheme over $\Fq$ and $G_m(\Fq) = \underline{G}(R_m)$.
Using Theorem~\ref{thm:geo}, let $\cs{L}_m$ be a geometrization of the character $\varphi_m: G_m(\Fq) \to \EEx$; so
\[
\trFrob{\cs{L}_m} = \varphi_m
\]
on $G_m(\Fq)$.
%
Recall that the full Greenberg transform $G \ceq \Gr^{R}(\underline{G})$ is a group scheme over $\Fq$ such that $G(\Fq) = \underline{G}(R)$; it comes equipped with a morphism $G \to G_m$.
Let $\cs{L}$ be the  Weil sheaf on $G$ obtained from $\cs{L}_m$ by pullback along $G \to G_m$. 
Then $\cs{L}$ is a quasicharacter sheaf on $G$, in the sense of \cite{cunningham-roe:13a}*{Def 4.3}, such that $\trFrob{\cs{L}} = \varphi$.
\end{proof}

>>>>

We are extremely grateful to Loren Spice for explaining Yu's types for supercuspidal representations and Masoud Kamgarpour for helpful conversations.
We also thank Will Sawin and the anonymous referee, who pointed out errors in earlier versions of this paper and helped us clarify several points.

@@

We now consider the case of a smooth group scheme without the commutativity assumption.  We start
by relating linear character sheaves on $G$ to character sheaves on its abelianization.

If $\chi \in G(k)^*$ is a character, it must vanish on the derived subgroup $G\der(k)$. \todo{Error} 
Kamgarpour gives an example \cite{kamgarpour:09a}*{(1.1)} of a linear character sheaf
on $G = \PGL_n$ that does not vanish on $G\der = \PGL_n$, defined by the extension
\[
1 \to \mu_n \to \SL_n \to \PGL_n \to 1
\]
together with a non-trivial character of $\mu_n(k)$. \todo{Small change here.}
In order to obtain a relationship between linear character sheaves on $G$ and characters of $G(k)$,
he opts to give a different definition of commutator and, in doing so, introduces a `stacky abelianization' of $G$ in order to geometrize characters of $G(\Fq)$.
 Since we have already seen the need to adapt the
notion of character sheaf, even in the commutative case, we instead add restrictions
to the definition of commutative character sheaf and leave the definition of $G\der$ unchanged, allowing us to use the schematic abelianization of $G$ in the geometrization of characters of $G(\Fq)$.

@@

Recall from Section~\ref{sec:defs} that we refer to objects in category $\CS(G)$, defined as
in \cite{cunningham-roe:13a}*{Def. 1.1}, as linear character sheaves when $G$ is smooth but
not necessarily commutative.  
We define the following category to track the trivialization on the derived subgroup;
commutative character sheaves will then be defined as a subcategory.

@@

The reason for tracking $\beta$ is that it determines an isomorphism $\gamma : m^*\cs{L} \to \theta^*m^*\cs{L}$ as follows,
which will replace the $\xi$ of Definition \ref{def:CCScom}.
Let $i : G \to G$ be inversion and $c : G\times G\to G\der$ be the commutator map, defined by $c(x,y)= xyx^{-1}y^{-1}$.
Both are morphisms of $\Fq$-schemes.
Set $m' = i \circ m \circ \theta$ and let $i\der : G\der\to G$ be inclusion; then $i\der\circ c = m \circ (m \times m')$.
Then $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ by the diagram of isomorphisms below.
\begin{equation}
\begin{tikzcd}
\arrow[equal]{d} c^* (\cs{L}\vert_{G\der}) \arrow{r}{c^*(\beta)} 
	&  c^*((\EE)_{G\der}) \arrow[equal]{d} \\
\arrow{d}[swap]{i\der\circ c = m \circ (m \times m')} c^* i\der^* \cs{L} 
	&   (\EE)_{G\times G} \\
(m \times m')^* m^* \cs{L} \arrow{d}[swap]{(m \times m')^*(\mu)} 
	&  m^*\cs{L} \otimes \theta^* m^* i^* \cs{L}  \arrow[dashed]{u}[swap]{\gamma'} \\
(m \times m')^* (\cs{L} \boxtimes \cs{L}) \arrow[equal]{r} 
	& m^*\cs{L} \otimes (m')^*\cs{L} \arrow{u}[swap]{m' = i\circ m\circ \theta} 
\end{tikzcd}
\end{equation}
In the diagram above, the arrows labeled with equations come from canonical isomorphisms of functors on Weil sheaves derived from the equations; so, for example, the middle left isomorphism comes from $(m\times m')^* m^* \iso c^* i\der^*$ since $i\der\circ c = m \circ (m \times m')$.
Using the monoidal structure of the category of Weil local systems on $G\times G$, the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ defines an isomorphism
\[
m^*\cs{L} \to (\theta^* m^* i^*\cs{L})^\vee.
\]
Applying the canonical isomorphisms $(\theta^* m^* i^*\cs{L})^\vee \iso \theta^* m^* i^* (\cs{L}^\vee)$ and $i^*(\cs{L}^\vee) \cong \cs{L}$, this map provides the promised isomorphism
\[
\begin{tikzcd}
\gamma : m^*\cs{L} \arrow{r} & \theta^* m^* \cs{L}.
\end{tikzcd}
\]

@@

\begin{lemma}\label{lem:HH}
If $(\cs{L},\mu,\nu)$ is a $G\der$-equivariant linear character sheaf on $G$ then $\mu : m^*\cs{L} \to \cs{L}\boxtimes \cs{L}$ 
is a morphism of $G^2\der$-equivariant Weil local systems on $G^2$, as defined in Section \ref{ssec:equivariant1}.
\end{lemma}

@@

\begin{corollary}
If $G$ is a smooth group scheme and $(\cs{L},\mu)$ is a linear character sheaf on $G$ with trivial restriction to $G\der$,
then the set of isomorphism classes of objects in $\CS(G\ab)$ mapping to $(\cs{L},\mu)$ under $q^*$ is a principal homogeneous space
for $C$.
\end{corollary}
\begin{proof}
By Proposition~\ref{prop:CSGabab}, it suffices to find the set of isomorphism classes in $\CSab(G)$ mapping to
$(\cs{L},\mu)$ under the forgetful functor.  By the previous corollary this set is nonempty.
If $(\cs{L},\mu,\beta)$ and $(\cs{L},\mu,\beta')$ both map to $(\cs{L},\mu)$, then $\beta' \circ \beta^{-1}$ is an
automorphism of the constant sheaf on $G\der$.  Conversely, if $\varphi$ is an automorphism of $(\EE)_{G\der}$
and $(\cs{L},\mu,\beta) \in \CSab(G)$ then $(\cs{L},\mu,\varphi \circ \beta) \in \CSab(G)$.  As in \cite{cunningham-roe:13a}*{Theorem 3.9},
the automorphism group is isomorphic to $\Hom(\pi_0(\bar{G}\der)_{\Frob{}}, \EEx)$.  Finally, we note that any
automorphism $\alpha$ of $(\cs{L},\mu) \in \CS(G)$ defines an isomorphism $(\cs{L},\mu,\beta \circ \alpha|_{G\der}) \to (\cs{L},\mu,\beta)$.
Applying the analogue of \cite{cunningham-roe:13a}*{Theorem 3.9} again yields the desired result.
\end{proof}

@@

\begin{lemma}\label{lemma:morinCSab}
If $(\cs{L},\mu,\beta)\in \CS\ab(G)$ then $\mu  : m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$, $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are morphisms in $\CS\ab(G^2)$.
\end{lemma}
\begin{proof}
Define $m^2 : G^2\times G^2 \to G^2$ by $m^2(g_1,g_2,g_1',g_2') = (g_1g_1',g_2g_2')$. Also define $p^2_i : G^2\times G^2 \to G^2$ by $p^2_i(g_1,g_2,g_1,'g_2') = (g_i,g_i')$.
First we show that $m^*\cs{L}$ is an object in $\CS(G^2)$ by equipping it with an isomorphism $\mu_m^2: (m^2)^* (m^*\cs{L}) \to m^*\cs{L} \boxtimes m^*\cs{L}$ defined by the diagram below.
\[
\begin{tikzcd}
(m^2)^* (m^*\cs{L}) \arrow{d}{(m^2)^*\mu} \arrow[dashed]{rr}{\mu_m^2} && m^*\cs{L} \boxtimes m^*\cs{L}\\
 (m^2)^*(\cs{L}\boxtimes\cs{L}) \arrow[equal]{r} & (m^2)^*(p_1)^*\cs{L} \otimes (m^2)^*(p_2)^*\cs{L} \arrow{r} & (p^2_1)^*m^*\cs{L} \otimes (p^2_2)^*m^*\cs{L} \arrow[equal]{u}
\end{tikzcd}
\] 
The pair $(m^*\cs{L},\mu_m^2)$ satisfies the conditions appearing in \cite{cunningham-roe:13a}*{Def. 1.1}.
The restriction of $m^*\cs{L}$ to $G^2\der= G\der\times G\der$ is canonically isomorphic to $(\EE)_{G^2\der}$ by
\[
\begin{tikzcd}
(m^*\cs{L})\vert_{G^2\der} \arrow{d}{\mu\vert_{G^2\der}} \arrow[dashed]{r}{\beta^2_m} & (\EE)_{G^2\der}\\
(\cs{L}\boxtimes\cs{L})\vert_{G^2\der} \arrow[equal]{r} & (\cs{L}\vert_{G\der})\boxtimes(\cs{L}\vert_{G\der}) \arrow{u}{\beta\boxtimes\beta}.
\end{tikzcd}
\]
This shows that $(m^*\cs{L},\mu_m^2,\beta_m^2) \in \CS\ab(G^2)$. 
Similar work defines $(\cs{L}\boxtimes\cs{L},\mu_\boxtimes^2,\beta_\boxtimes^2) \in \CS\ab(G^2)$.
By construction, $\mu: m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$ is a morphism in $\CS\ab(G^2)$.
Similar work shows that $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are also morphisms in $\CS\ab(G^2)$.
\end{proof}

@@

\begin{proof}
By continuity of $\varphi : \underline{G}(R) \to \EE^\times$, there is some $m \in \NN$ and a factorization
\[
\begin{tikzcd}
\underline{G}(R) \arrow{rr}{\varphi} \arrow{rd} && \EEx\\
& \underline{G}(R/\mathfrak{p}^{m+1}) \arrow{ru}[swap]{\varphi_m} 
\end{tikzcd}
\] 
Set $R_m = R/\mathfrak{p}^{m+1}$ and
set $G_m = \Gr_m^{R}(\underline{G})$, the Greenberg transform of $\underline{G}\times_{\Spec{R}}\Spec{R_m}$.
Then $G_m$ is a smooth group scheme over $\Fq$ and $G_m(\Fq) = \underline{G}(R_m)$.
Using Theorem~\ref{thm:geo}, let $\cs{L}_m$ be a geometrization of the character $\varphi_m: G_m(\Fq) \to \EEx$; so
\[
\trFrob{\cs{L}_m} = \varphi_m
\]
on $G_m(\Fq)$.
%
Recall that the full Greenberg transform $G \ceq \Gr^{R}(\underline{G})$ is a group scheme over $\Fq$ such that $G(\Fq) = \underline{G}(R)$; it comes equipped with a morphism $G \to G_m$.
Let $\cs{L}$ be the Weil sheaf on $G$ obtained from $\cs{L}_m$ by pullback along $G \to G_m$. 
Then $\cs{L}$ is a quasicharacter sheaf on $G$, in the sense of \cite{cunningham-roe:13a}*{Def 4.2}, such that $\trFrob{\cs{L}} = \varphi$.
\end{proof}
